---
title: "Project2_StockAnalysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Project2_StockAnalysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rotateS21)
```

# Intro

Stock markets are made up of many different stocks. Trying to analyze all of the stocks at once can turn into a multi-dimensional nightmare. One possible solution is 'reducing' the dimensions of the problem with a PCA. With a PCA one can summarize the data using a series of orthogonal linear combinations that capture most of the variation in the data. We will recreate example 8.10 to go through an analysis of how a PCA can be used for stock market data collected for 5 different stocks over a span of 103 weeks. 

```{r}
stock <- read.table("../inst/df/T8-4.DAT")
names(stock) <- c("JP", "Citi", "Wells", "Royal", "Exxon")
head(stock)
```

We can use the eigenPCA function to perform most of the analysis.

# Part a

The first thing that we want to do is perform the PCA itself. This is done by finding the covariance matrix and analyzing the eigenvalues and eigenvectors. 



